Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-17 ebayes_rnaseq_heterosis.tex
changed notation

To identify genes displaying heterosis using RNAseq counts, we build a hierarchical model to borrow information on gene-variety means as well as overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.

@@

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$. As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$. Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.
\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\delta_{vi}},\psi_g) 
\label{e:data}
\end{equation}

A hierarchical model is constructed for the sequencing depth, overdispersion, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\delta_{vi} \stackrel{ind}{\sim} N(0,\tau_\delta^2)$. The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e. 
$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$. We consider an \emph{independent} model where $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model where we estimate all parameters in $\Sigma$.

We assume improper uniform priors for $\eta$ and $\tau_\delta\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution. In the independent model, we also $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.

@@

The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\phi = (\delta,\eta, \Sigma)$ where $\delta = (\delta_{11},\ldots,\delta_{Vn_V})$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\phi$, denoted $\hat{\phi}$, and then base gene-specific inference on the posterior conditional on this estimated hyperparameter. Conditional on the hyperparameters, the gene-specific parameters are conditionally independent as in equation \eqref{e:condind}. Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
\begin{align}
p\left(\theta\left|y,\hat{\phi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\phi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\delta}\right.\right) p\left(\theta_g\left|\hat{\phi}\right.\right) \nonumber \\
&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\delta}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right) 
\label{e:condind}
\end{align}

@@

As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate for $\theta$ and $\phi$. We explicitly use the MAP estimate for $\phi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}. The MAP estimate for $\theta$ can also be used in this step as initial values for the MCMC.

@@

As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
\label{e:hypotheses}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence for extreme heterosis. This list can be used by geneticists to prioritize future experiments to understand the genetic mechanism for heterosis.

@@

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\gamma_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\zeta_g = \mu_{g3} - \gamma_g$ be the hybrid effect. After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\zeta}_g$ and $\hat{\alpha}_g$ of $\zeta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$. Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\zeta}_g| \le |\hat{\alpha}_g|$, $P(\zeta_g + \alpha_g > 0 |y)$ if $\hat{\zeta}_g > -\hat{\alpha}_g \ge 0$, $P(\zeta_g - \alpha_g > 0 |y)$ if $\hat{\zeta}_g > \hat{\alpha}_g \ge 0$, $P(\zeta_g - \alpha_g < 0 |y)$ if $\hat{\zeta}_g < \hat{\alpha}_g \le 0$, and $P(\zeta_g + \alpha_g < 0 |y)$ if $\hat{\zeta}_g < -\hat{\alpha}_g \le 0$.

@@

Figure \ref{f:auc} provides the area under the ROC below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two. While there does not appear to be a significant difference between the independence model using RStan or ShrinkBayes, there does appear to be a significant decreased performance, at least for smaller sample sizes, when using the model where a covariance is estimated.

\section{Heterosis in maize ??}
\label{s:corn}

>>>>

To identify genes displaying heterosis using RNAseq counts, we build a hierarchical model to borrow information on gene-variety means and gene-specific overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.

@@

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$. As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$. 
\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g) 
\label{e:data}
\end{equation}
where $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.

A hierarchical model is constructed for the sequencing depth, overdispersion, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau_\gamma^2)$. (These terms are the normalization factors that are typically estimated from each sample and then plugged in for further analysis.) The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e. 
$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$. We consider an \emph{independent} model where $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model where we estimate all parameters in $\Sigma$.

We assume improper uniform priors for $\eta$ and $\tau_\gamma\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution. In the independent model, we also $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.

@@

The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\pi = (\gamma,\eta, \Sigma)$ where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this estimated hyperparameter. Conditional on the hyperparameters, the gene-specific parameters are conditionally independent as in equation \eqref{e:condind}. Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
\begin{align}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right) 
\label{e:condind}
\end{align}

@@

As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate for $\theta$ and $\pi$. We explicitly use the MAP estimate for $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}. The MAP estimate for $\theta$ can also be used in this step as initial values for the MCMC.

@@

As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
\label{e:hypotheses}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence for extreme heterosis. This list can be used by geneticists to prioritize future experiments to understand the genetic mechanism for heterosis.

@@

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$. Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

Figure \ref{f:auc} provides the area under the ROC below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two. While there does not appear to be a significant difference between the independence model using {\tt RStan} or {\tt ShrinkBayes}, there does appear to be a significant decreased performance, at least for smaller sample sizes, when using the model where a covariance is estimated.

\section{Heterosis in maize experiment}
\label{s:corn}
